---
title: "Introduction to coca"
author: "Alessandra Cabassi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: united
vignette: >
  %\VignetteIndexEntry{R package klic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The R package **coca** contains the functions needed to use COCA (**C**luster-**O**f-**C**lusters **A**nalysis), an integrative clustering method that was first introduced in a breast cancer study by The Cancer Genome Atlas in 2012 and quickly became a popular tool in cancer studies (see e.g. Hoadley et al. 2014 and Aure et al. 2017). It is based on \emph{Consensus Clustering} (Monti et al., 2013), an algorithm that was initially developed to assess the stability of the clusters obtained with any clustering algorithm. 

# Building the matrix of clusters

The main goal of COCA is to summarise the clusterings found in different ’omics datasets, by identifying a “global" clustering across the datasets that is intended to be in good agreement with the clustering structures
identified in each of the individual datasets.

The first step of COCA is the construction of the Matrix-Of-Clusters (MOC). This is a binary matrix of size _N x K_, where _K_ is the sum of the number of clusters _Km_ in every dataset _Xm_. Therefore, to each column _j_ of this matrix corresponds a cluster _mk_ in dataset _Xm_. The _(i,j)_-th entry of the matrix of clusters is equal to one if data point _i_ belongs to cluster _mk_ in dataset _Xm_, and is equal to zero otherwise.For more details about how the matrix of clusters is built, please see Cabassi and Kirk (2018).

In the example below, we assume to have three datasets with the same number of clusters, six. The clustering structure in each of those dataasets is found via k-means clustering. The resulting clusters are stored in three matrices of cluster labels, that are then collated together to get a matrix of clusters.

```{r preparation_coca, fig.show='hold', message=FALSE, warning=FALSE, cache=TRUE}
### Load data
data <- list()
data[[1]] <- as.matrix(read.csv(system.file("extdata", 
                      "dataset1.csv", package = "coca"), row.names = 1))
data[[2]] <- as.matrix(read.csv(system.file("extdata", 
                      "dataset2.csv", package = "coca"), row.names = 1))
data[[3]] <- as.matrix(read.csv(system.file("extdata", 
                      "dataset3.csv", package = "coca"), row.names = 1))

### Set constants
n_datasets <- 3
n_clusters <- 6
N <- 300
true_labels <- as.matrix(read.csv(system.file("extdata",
                         "cluster_labels.csv", package = "coca"), row.names = 1))

### Fill label matrix with clusterings found with the k-means clustering algorithm
labelMatrix <- array(NA, c(n_clusters, N, n_datasets))
for(i in 1:n_datasets){
  output <- kmeans(data[[i]], n_clusters)
  for(k in 1:n_clusters){
    labelMatrix[k,,i] <- (output$cluster==k)
  }
}

# Convert label matrix from logic to numeric matrix
labelMatrix <- labelMatrix*1

# Build MOC matrix
MOC <- rbind(labelMatrix[,,1],labelMatrix[,,2], labelMatrix[,,3])
```

# Cluster-Of-Clusters Analysis

The MOC matrix is then used as input to consensus clustering (CC), an algotithm that was developed by Monti et al. (2003) to assess cluster stability when analysing a single dataset. The resulting consensus matrix is then used as
the similarity matrix for a hierarchical clustering method (or any other distance-based clustering algorithm). These last two steps of the COCA algorithm are contained in the `coca()` function. 

```{r coca, fig.show='hold', message=FALSE, warning=FALSE, cache=TRUE}
### COCA

# Use COCA to find global clustering
coca <- coca(t(MOC), K = 6, hclustMethod = "average")

# Compare clustering to the true labels
ari <- mclust::adjustedRandIndex(true_labels, coca$clusterLabels)
ari
```

# References 

Aure, M.R., Vitelli, V., Jernström, S., Kumar, S., Krohn, M., Due, E.U., Haukaas, T.H., Leivonen, S.K., Vollan, H.K.M., Lüders, T. and Rødland, E. (2017). Integrative clustering reveals a novel split in the luminal A subtype of breast cancer with impact on outcome. Breast Cancer Research, 19(1), p.44.

Cabassi, A. and Kirk, P. D. W. (2018). Multiple kernel learning for integrative consensus clustering. In preparation.

Hoadley, K.A., Yau, C., Wolf, D.M., Cherniack, A.D., Tamborero, D., Ng, S., Leiserson, M.D., Niu, B., McLellan, M.D., Uzunangelov, V. and Zhang, J., 2014. Multiplatform analysis of 12 cancer types reveals molecular classification within and across tissues of origin. Cell, 158(4), pp.929-944.

Monti, S. et al. (2003). Consensus Clustering: A Resampling-Based Method for Class Discovery and Visualization of Gene. Machine Learning, 52(i), 91–118.

The Cancer Genome Atlas (2012). Comprehensive molecular portraits of human breast tumours. Nature,
487(7407), 61–70.
